The Name of the Title Is Hope

### 2024-12-30 paper.tex
working through the text, wip

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{The Scan Primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. \john{Rabin-Karp string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, and is a good thing to cite in a paper going to SPAA\@.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Prior to the introduction of shared memory, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory and performed a single global scan where each level of depth necessitated a separate kernel launch. The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies. \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory. However, as a natural result of this partitioning, a serial dependency is created between the work tiles. In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.

>>>>

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
\john{So I have written, roughly, a zillion of these sections and here is my current philosophy. The danger is that you write something really interesting and relevant and a reviewer who is familiar with GPUs sees the title and sees the first few lines and says ``I can skip this'' and does. There are plenty of great descriptions of how GPUs work. Cite one of them. What this section needs to do is very specifically call out the interesting/detailed features of GPUs that are \emph{particularly relevant to this paper} that are not just ``here's generally how a GPU works''. For me, that's the last paragraph. The other part that might be necessary is defining the specific terminology used in the paper (e.g., workgroup vs.\ thread block vs.\ CTA). If you feel like breaking up a wall of text, which I generally think is a good idea, we can make a little table of equivalent terminology and boldface the terms we're using.} \john{Perhaps another good thing to mention that is specific to this paper is the GPU's support for synchronization. I could see having a blurb on ``there is no global inter-workgroup synchronization except an expensive kernel barrier'' and also what \emph{is} available in terms of pairwise synchronization between workgroups and what the requirements are (e.g., residence). These are more detailed (and important) GPU topics that aren't present in the higher-level summary you have below. Later in the paper, you reference ``without an inter-workgroup barrier primitive'' and don't say why, and you don't want the reader to get the wrong idea here (you don't want the reader to say ``oh, we can just add this to the language, we're done''). There is a fundamental reason to not have this primitive. Here's where you can discuss that.} \john{Thinking about the above, one way to organize this subsection that is most helpful is to \emph{show} here what kind of synchronizations are / aren't possible within the GPU programming model, in a principled way, and then these are the constraints for our implementation. We want readers to not just accept a weird set of constraints as a given but rather to understand where they come from, and why.}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

\subsection{The Scan Primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{10.1145/322217.322232, 5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. \john{Rabin-Karp's second fingerprinting method for string matching (``Efficient randomized pattern-matching algorithms'', Section 6, 1987)~\cite[Section 6]{Karp:1987:ERP} has 2$\times$2 matrix multiplication as its binary operation, which is non-commutative, and is a good thing to cite in a paper going to a theory conference like SPAA, because string matching is a super-SPAA-friendly topic. I put it in the bib if you want to use it.} In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris offers a taxonomy that relates depth, fanout, and wire tracks~\cite{}, while Hinze develops an algebraic framework for scans~\cite{}. Snir proved that \emph{depth} and \emph{size} are related by (fill in)~\cite{}, and Fich proved that among minimum depth scans Ladner-Fischer networks have optimal size. Finally, Merrill and Garland provide a comprehensive review of GPU scan implementations. In this work, we focus on the evolution of inter-workgroup (global) scan strategies and evaluate their suitability for our ideal architecture.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Prior to the introduction of shared memory\john{pre-CUDA?}, early GPU scan implementations~\cite{Horn, Hensley, Greß, Sengupta} operated directly on global memory \john{the previous implementation strategy is PRAM-like and you should say that, it's important to a SPAA audience} and performed a single global scan where each level of depth necessitated a separate kernel launch\john{this latter phrase is not obvious! explain why}. The introduction of shared memory and dedicated compute environments resulted in the first implementations to resemble contemporary designs, combining coarse-grained inter-workgroup strategies with fine-grained intra-workgroup strategies. \emph{Work tile} partitioning was introduced to divide inputs into workgroup-consumable chunks that fit within the limited size of shared memory. However, as a natural result of this partitioning, a serial dependency is created between the work tiles. In this section, we review the three prominent global scan strategies—\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}—examining how they navigate this inter-workgroup dependency and evaluating their effectiveness in minimizing global memory access and maximizing memory bandwidth utilization in alignment with our ideal architecture goals.
